Normalizations are not what you think; Explicit Scale Simulation in ALDEx2

### 2024-08-13 go3.tex
go3-edits
final go3 before Justin

\begin{document}
\maketitle
\begin{abstract}
In many high-throughput sequencing (HTS) studies, sample-to-sample
variation in sequencing depth is not driven by variation in the scale
(e.g., total size, microbial load, or total gene expression) of the
underlying biological systems being measured but rather is driven by
technical factors. Typically the technical variation is addressed using
some form of statistical normalization. Normalizations are data or
parameter transformations that remove unwanted technical variation in
the hopes of facilitating analyses sensitive to scale; e.g.,
differential abundance and differential expression analyses. Recently we
showed that any normalization makes implicit assumptions about the
unmeasured system scale and that errors in these assumptions can lead to
dramatic increases in false positive and false negative rates. Here we
describe updates to the ALDEx2 R package that mitigate these problems by
directly modeling uncertainty in the unmeasured system scale through the
use of a \textit{scale model}. Scale models generalize the idea of
normalizations and can be thought of as explicitly modeling potential
error in the chosen normalization. Beyond enhancing the robustness of
HTS analyses, the use of scale models within ALDEx2 enhances the
transparency and reproducibility of analyses by making implicit
normalizing assumptions an explicit part of the model building process.
\end{abstract}

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

High-throughput sequencing (HTS) is a ubiquitous tool used to explore
many biological phenomenon such as gene expression (single-cell
sequencing, RNA-sequencing, meta-transcriptomics), microbial community
composition (16S rRNA gene sequencing, shotgun metagenomics) and
differential enzyme activity (selex, CRISPR killing). HTS proceeds by
taking a sample from the environment, making a library, multiplexing
(merging) multiple libraries together, and then applying a sample of the
multiplexed library to the flow cell. Each of these steps can be though
of as a sampling step: a pool of DNA that could be measured is sampled
and only that subsampled DNA is carried over to subsequent steps. Within
each sampling step, the connection between the size of the sampled DNA
pool and the scale (e.g., size, microbial load, or total gene
expression) of the measured biological system is degraded or lost.
Increasingly, researchers are turning to modified experimental protocols
(e.g., DNA spike-ins or cell counting) in an attempt to recover the lost
biological variation in scale; see for example (Vandeputte et al. 2017;
Props et al. 2017). However, these protocols often do not recover
meaningful biological variation but instead recover scale variation at
some intermediate step in the sample preparation protocol (e.g., at the
stage where the DNA spike-in was added). In short, the disconnect
between sample-to-sample variation in sequencing depth and biological
variation in scale remains an outstanding challenge.

Biological variation in scale can represent an important unmeasured
confounder in HTS analyses (Lovell et al. 2015). For example, cells
transformed by the cMyc oncogene have about 3 times the amount of mRNA
and about twice the rRNA content than non-transformed cells (Nie et al.
2012), and this dramatically skews transcriptome analysis (Lovén et al.
2012). In addition, wild-type and mutant strains of cell lines, yeast or
bacteria often have different growth rates, which would affect our
ability to identify truly differentially abundant genes (Yoshikawa et
al. 2011). As another example, the total bacterial load of the vaginal
microbiome differs by 1-2 orders of magnitude in absolute abundance
between the healthy and bacterial vaginosis states (Zozaya-Hinchliffe et
al. 2010), and the composition between these states is dramatically
different (Ravel et al. 2011; Hummelen et al. 2010). Thus, a full
description of these systems includes both relative change (composition)
and absolute abundance (scale), yet current methods access only the
compositional information.

To be more concrete, let \(\mathbf{Y}\) denote an \(D \times N\) matrix
of sequence counts with elements \(\mathbf{Y}_{dn}\) denoting the number
of measured DNA molecules mapping to feature \(d\) (e.g., a taxon or
gene) in sample \(n\). Let \(\mathbf{W}\) also represent a
\(D \times N\) matrix but with elements \(\mathbf{W}_{dn}\) representing
the true (as opposed to measured) amount of class \(d\) in the
biological system from which sample \(n\) was obtained. Succinctly,
\(\mathbf{Y}\) denotes the observed HTS data whereas \(\mathbf{W}\)
denotes the true counts in the system under study. We can think of
\(\mathbf{W}\) as consisting of two parts, the scale
\(\mathbf{W}^{\perp}\) (e.g., totals) and the composition
\(\mathbf{W}^{\parallel}\) (i.e., proportions). That is,
\(\mathbf{W}^{\perp}\) is a \(N\)-vector with elements
\(\mathbf{W}^{\perp}_{n}=\sum_{d}\mathbf{W}_{dn}\) while
\(\mathbf{W}^{\parallel}\) is a \(D \times N\) matrix with elements
\(\mathbf{W}^{\parallel}_{dn}=\mathbf{W}_{dn}/\mathbf{W}^{\perp}_{n}\).
Note that with these definitions \(\mathbf{W}\) can be written as the
element-wise combination of scale and composition:
\(\mathbf{W}_{dn}=\mathbf{W}^{\parallel}_{dn}\mathbf{W}^{\perp}_{n}\).
The technical variation in sequencing depth
(\(\mathbf{Y}^{\perp}_{n}=\sum_{d}\mathbf{Y}_{dn}\)) implies that
observed data \(\mathbf{Y}\) provides us with information about the
system composition \(\mathbf{W}^{\parallel}\) but little to no
information in the system scale \(\mathbf{W}^{\perp}\) (Lovell et
al.~2011).

Early on, researchers recognized the potential confounding effects of
the non-biological variation in sequencing depth and approached the
problem using various forms of \textit{normalization}. The idea of
normalization is to remove this variation thus making the samples
commensurate and to correct any minor asymmetries in the data. A large
number of normalizations were developed. Rarefaction (Hughes and
Hellmann 2005) was used in the microbiome field, while transcriptome
investigators developed and used proportions and derivatives (reads per
kilobase per million, and transcripts per kilobase per million)
(Mortazavi et al. 2008; Wagner, Kin, and Lynch 2012), the relative log
expression (RLE) (Anders and Huber 2010) and trimmed mean of M values
(TMM) (Robinson and Oshlack 2010). The Centered Log Ratio (CLR)
(Aitchison 1982) was proposed as a unifying normalization (Fernandes et
al. 2014). With the exception of rarefaction, all of these
normalizations are data transformations that can be stated as ratios of
the form \(\hat{\mathbf{Y}}_{dn}=\mathbf{Y}_{dn}/f(\mathbf{Y})\); the
major differences being how the denominator is chosen and whether the
ratio is subsequently always log-transformed or not.

Recently, Nixon et al. (2023) suggested that the challenge of
non-biological variation in sequencing depth be viewed as a problem of
partially-identified models. Through this lens, they showed that
\emph{all} normalizations make some assumption about the system scale
\(\mathbf{W}^{\perp}\) in order to identify the chosen model. Moreover,
these assumptions are often implicit, difficult to interpret, and
provided different outputs when applied to the same dataset (Dillies et
al. 2013; Weiss et al. 2017). Intuitively, normalizations in widespread
use assume that either one sample can be chosen as a reference to which
the others are scaled (TMM), or that different sub-parts of each sample
maintain a constant scale across samples (RLE, CLR and derivatives).
Further, Nixon et al. (2023) showed that the original ALDEx2 model made
a strict assumption about scale through the CLR normalization. They
showed that this assumption could be remedied since ALDEx2 permits scale
to be easily incorporated into its framework. In this paper, we discuss
these modifications and show how including scale uncertainty can greatly
improve modeling and lead to more robust and reproducible results.

\hypertarget{adding-scale-uncertainty-in-aldex2}{%
\subsection{Adding Scale Uncertainty in
ALDEx2}\label{adding-scale-uncertainty-in-aldex2}}

The ALDEx2 R package (Fernandes et al. 2013) is a general purpose
toolbox for Bayesian modeling of HTS data. For brevity and concreteness,
we discuss ALDEx2 in its simplest form as a tool for estimating
Log-Fold-Changes (LFC) (e.g., differential abundance or differential
expression analysis), but note that it can be used to fit more complex
linear models. At a high-level, ALDEx2 involves three steps: 1) a
Bayesian model is used to estimate \(\mathbf{W}^{\parallel}\) given
observations \(\mathbf{Y}\); 2) a normalization is used to estimate
\(\mathbf{W}\) given the estimate \(\mathbf{W}^{\parallel}\); 3) the
\(\mathbf{W}\) estimates are used to estimate LFC as part of
differential abundance (or expression) analyses. For more details on
ALDEx2 see Fernandes et al. (2013).

By default, ALDEx2 uses the CLR normalization which can be written as
\(\log \mathbf{W}_{dn}=\log [\mathbf{W}^{\parallel}_{dn}/G_{n}]\) where
\(G_{n}\) denotes the geometric mean of the vector
\((W^{\parallel}_{1n}, \dots, W^{\parallel}_{Dn})\) (see Supplemental
Materials for how the CLR connects to Shannon entropy). Nixon et
al.~(2023) showed that in this context the CLR normalization equates to
an implicit assumption that \(\mathbf{W}^{\perp}_{n}=1/G_{n}\). They
showed that even slight errors in this assumption introduced bias into
LFC estimate that was not accounted for in estimates of uncertainty
(e.g., confidence intervals or credible sets). In fact, they showed that
the only way in which the ALDEx2 model, or any model that uses
normalizations, could ever be calibrated (e.g., control Type-I Error
rates) was if this assumption was exactly true. While this assumption
may be approximately correct in some instances it is surely never
exactly true. In support of this assertion it is known that false
discovery rates vary widely by analysis and normalization method
(Hawinkel et al. 2018; Li et al. 2022). As a result, Nixon et al.~(2023)
suggested that models should consider potential error in the assumptions
implied by normalizations.

Nixon et al.~(2023) proposed the concept of a \textit{scale model} as a
generalization of normalizations that account for potential error. They
showed how scale models can be incorporated into ALDEx2, turning the
ALDEx2 model into a specialized type of statistical model which they
called a \textit{Scale Simulation Random   Variable} (SSRV). Recall that
\(\mathbf{W}_{dn}=\mathbf{W}^{\parallel}_{dn}\mathbf{W}^{\perp}_{n}\).
As the HTS data only informs directly on
\(\mathbf{W}^{\parallel}_{dn}\), they proposed to augment analyses with
a model for \(\mathbf{W}^{\perp}_{n}\) (the scale model). Since the CLR
normalization makes the assumption \(\mathbf{W}^{\perp}_{n}=1/G_{n}\),
the CLR normalization can be generalized by considering probability
models for the scale \(\mathbf{W}^{\perp}_{n}\) that have mean
\(1/G_{n}\). For example, the following scale model generalizes the CLR:

\[\log \mathbf{W}^{\perp}_{n} = -\log G_{n} + \Lambda x_{n} \qquad \Lambda \sim N(0, \gamma^{2})\]

where \(\gamma\) is a tunable parameter that controls the degree of
uncertainty in the CLR assumption and \(x_{n}\) denotes a binary
condition indicator (e.g., \(x_{n}=1\) denotes case and \(x_{n}=0\)
denotes control). Beyond theoretical studies, they demonstrated through
the analysis of simulated and real data the remarkable performance
improvements that could result from incorporating this type of scale
uncertainty into analyses. Subsequently, we have made those
modifications a permanent fixture of ALDEx2 which now represents the
first software package designed for SSRV-based inference.

Beyond concerns of fidelity and rigor, scale models also enhance the
reproducibility and transparency of HTS analyses. This occurs because
rather than using an implicit assumption (e.g.,
\(\log \mathbf{W}^{\perp}=1/G_{n}\) for the CLR normalization), scale
models make this an explicit part of the model building process. While
it is beyond the scope of the present article, we note that there are
many ways of building scale models that enhance the interpretability of
the parameters and assumptions. Moreover, the scale model listed above,
which is now the default in ALDEx2, also has an appealing
interpretation. A detailed description of these points as well as the
mathematical underpinnings of the updated ALDEx2 R package is
forthcoming and will be made available as a companion publication. Here,
we simply treat the parameter \(\gamma\) in the default scale model as
an abstract term that represents uncertainty in the CLR assumption and
note that larger values of \(\gamma\) correspond to more uncertainty in
this assumption. We also show that the ratio of scales between
conditions can be used to build a full scale model when the conditions
have dramatically different underlying scales.

An advantage of incorporating scale is that analyses can be made much
more robust such that actual or potential differences in scale can be
tested and accounted for explicitly. The examples below and in the
supplement show how incorporating scale provides robust and
interpretable differential abundance estimates in several different
datasets.

\hypertarget{results}{%
\section{Results}\label{results}}

\hypertarget{adding-scale-uncertainty-replaces-the-need-for-dual-significance-cutoffs.}{%
\subsection{Adding scale uncertainty replaces the need for dual
significance
cutoffs.}\label{adding-scale-uncertainty-replaces-the-need-for-dual-significance-cutoffs.}}

Gierliński et al. (2015) conducted a highly replicated yeast
transcriptome experiment comparing a wild-type strain with a snf1 gene
knockout. This dataset has been used to argue that only a small number
of replicates are needed to identity differentially abundant genes and
makes the recommendation that some tools should be used in some
conditions because each tool has different intrinsic statistical power
and Type 1 and Type 2 errors (Schurch et al. 2016){]}. This guidance
runs counter to standard statistical practice where power is
intrinsically linked to sample size (Halsey et al. 2015), yet the
concept of sample-size independent power is entrenched in all fields
that use HTS as an experimental readout. Through the lens of this paper,
this is a hallmark of unacknowledged bias: a false certainty in the
scale of the data brings about an increasing false precision with
increasing sample sizes. Furthermore, the yeast transcriptome experiment
has an unacknowledged difference in scale between conditions: yeast
deficient for snf1 are smaller, grow more slowly and are sensitive to a
variety of common agents that cause cell stress (Yoshikawa et al.
2011){]}. Thus, by adding scale uncertainty, we can be more confident
that the results of the analysis reflect true biological differences and
not differences due to the assumptions made by the normalization method.

We start by examining the dispersion (variance) of the data as counts
and as the logarithm of the normalized counts when calculated by DESeq2
(log(RLE)) or by ALDEx2 (CLR). The raw data counts derived from
sequencing are overdispersed with the mean value being less than the
variance (Robinson and Oshlack 2010; Robinson, McCarthy, and Smyth 2010)
as seen in Panel A of Figure 1. However, the actual analysis of
differential abundance is performed on the logarithm of the normalized
counts (Robinson, McCarthy, and Smyth 2010; Love, Huber, and Anders
2014; Andrews and Hemberg 2019), which includes the CLR (Fernandes et
al. 2013) yet the mean-dispersion distribution of these log-ratio
transformed data is quite different as shown in panels B and C of Figure
1 and as noted elsewhere (Fernandes et al. 2013; Love, Huber, and Anders
2014).

\begin{figure}
\centering
\includegraphics{go3_files/figure-latex/dispersion-1.pdf}
\caption{Plot of abundance v dispersion for a typical transcriptome
dataset as counts, as logarithms of counts, and as clr values. Panel A
shows that the data are over-dispersed relative to a Poisson
distribution which is represented by the dashed line when plotted on a
log-log scale. Panel B shows that the relationship between the mean and
the dispersion calculated in DESeq2, here the standard error (SE) of the
mean, is very different when the data are log-transformed first. Panel C
shows the equivalent values calculated by ALDEx2 in which the expected
clr value for each transcipt are plotted vs.~the expected dispersion.
The red line in each panel shows the loess line of fit to the mid-point
of the distributions. In panels B and C the amount of dispersion reaches
a minimum at moderate values. The dashed orange line in panel A is the
line of equivalence, and in panel B and C is the minimum y value. The
values below the dashed grey line in panels B and C represent those
below the first decile of dispersion.}
\end{figure}

These relationships depend on what is being analyzed: the variance or
dispersion always increases with increasing raw (or normalzied) read
count but decreases when measured on the log-ratio transformed data
(Fernandes et al. 2013; Love, Huber, and Anders 2014), reaching a
minimum at some mid-point of the distribution. This makes the
counter-intuitive suggestion that genes with moderate expression have
more predictable expression than genes with very high expression such as
housekeeping genes. This is at odds with the known biology of cells
where single cell counting of housekeeping transcripts shows that they
are both highly expressed and have little intrinsic variation (Taniguchi
et al. 2010). Furthermore, the dispersion is exceeding small being, for
many transcripts, almost negligible. To show this point more clearly,
the majority of the transcripts in the lowest decile of dispersion
indicated below the dashed grey line are statistically significantly
different (75\% with DESeq2, 69\% with ALDEx2), suggesting that low
dispersion estimates lead to many false positives. From the perspective
of Nixon et al. (2023), it is reasonable to conclude that some of these
results may be due to the choice of normalization. Indeed, benchmarking
and comparison studies have repeatedly shown that the choice of
normalization plays a role in which results are returned as significant;
see for example (Maza et al. 2013), (Dillies et al. 2013) and (Weiss et
al. 2017).

Using either DESeq2 or ALDEx2, a majority of transcripts are
statistically significantly different between groups even with a
Benjamini-Hochberg (Benjamini and Hochberg 1995) false discovery rate
(FDR) of 0.01; i.e.~4264 or 3790 of the 5891 transcripts are
statistically significant. Clearly, a 70\% positive rate seems
biologically unjustified, and furthermore this breaks the necessary
assumption made by the normalization methods that most features must be
invariant. Finally, that 90 transcripts are identified by ALDEx2 and not
DESeq2, while DESeq2 identifies 564 transcripts that ALDEx2 does not
suggests that the choice of normalization plays a role in which results
are returned as significant.

>>>>

\begin{document}
\maketitle
\begin{abstract}
In high-throughput sequencing (HTS) studies, sample-to-sample variation
in sequencing depth is driven by technical factors, and not by variation
in the scale (e.g., total size, microbial load, or total gene
expression) of the underlying biological systems. Typically a
statistical normalization is used to remove unwanted technical variation
in the data or the parameters of the model to enable analyses that are
sensitive to scale; e.g., differential abundance and differential
expression analyses. Recently we showed that all normalizations make
implicit assumptions about the unmeasured system scale and that errors
in these assumptions can lead to dramatic increases in false positive
and false negative rates. Here we describe updates to the ALDEx2 R
package that mitigate these problems by directly modeling uncertainty in
the unmeasured system scale through the use of a \textit{scale model}.
Scale models generalize the idea of normalizations and can be thought of
as explicitly modeling the error in normalization by examining a
distribution over all possible normalizations. Beyond enhancing the
robustness of HTS analyses, the use of scale models within ALDEx2
enhances the transparency and reproducibility of analyses by making
implicit normalizing assumptions an explicit part of the model building
process.
\end{abstract}

\section{Introduction}\label{introduction}

High-throughput sequencing (HTS) is a ubiquitous tool used to explore
many biological phenomenon such as gene expression (single-cell
sequencing, RNA-sequencing, meta-transcriptomics), microbial community
composition (16S rRNA gene sequencing, shotgun metagenomics) and
differential enzyme activity (selex, CRISPR killing). HTS proceeds by
taking a sample from the environment, making a library, multiplexing
(merging) multiple libraries together, and then applying a sample of the
multiplexed library to the flow cell. Each of these steps is a
compositional sampling step as only a fixed-size subsample of nucleic
acid is carried over to subsequent steps. Thus, with each sampling step
the connection between the size of the sampled DNA pool and the scale
(e.g., size, microbial load, or total gene expression) of the measured
biological system is degraded or lost. Increasingly, researchers are
turning to modified experimental protocols (e.g., DNA spike-ins or cell
counting) in an attempt to recover the lost biological variation in
scale; see for example (1, 2). However, these protocols often do not
recover meaningful biological variation but instead recover scale
variation at the intermediate step in the sample preparation protocol
where the DNA spike-in was added. In short, the disconnect between
sample-to-sample variation in sequencing depth and biological variation
in scale remains an outstanding challenge.

The analysis of HTS data suffers from several known problems that can be
traced, in whole or in part, to misspecification of scale. The first
issue is poor control of the false discovery rate (FDR)
{[}(3);(4);Nearing:\href{mailto:2022aa;@Li}{\nolinkurl{2022aa;@Li}}:2022aa{]},
exhibited as dataset-dependent FDR control. The FDR problem is connected
to the double-filtering approach that is often used, but which is known
not to have appropriate FDR control (5, 6). The second issue is poor
performance when analyzing asymmetric data; data that arises because of
a bias in the direction of change. This type of data frequently arises
in in-vitro selection experiments (SELEX), transcriptome analysis, and
microbiome analysis (7). The third issue is that these problems become
more pronounced as more samples are collected; that is, more information
results in a worsening of the accuracy of the analysis (9). The final
problem is that these data are relative data, i.e.~compositional (10,
11), and substantial effort has been put into removing this constraint.

The first three problems were recently shown by Nixon and Silverman (8)
to be a result of a mismatch between the underlying size or scale of the
system and the assumptions of the normalizations used for the analysis
of HTS. Biological variation in scale often represents an important
unmeasured confounder in HTS analyses (12). For example, cells
transformed by the cMyc oncogene have about 3 times the amount of mRNA
and about twice the rRNA content than non-transformed cells (13), and
this dramatically skews transcriptome analysis (14). In addition,
wild-type and mutant strains of cell lines, yeast or bacteria have
different growth rates and RNA contents under different conditions,
which affect our ability to identify truly differentially abundant genes
(15--17). As another example, the total bacterial load of the vaginal
microbiome differs by 1-2 orders of magnitude in absolute abundance
between the healthy and bacterial vaginosis states (18), and the
composition between these states is dramatically different (19, 20).
Thus, a full description of any of these systems includes both relative
change (composition) and absolute abundance (scale). Current methods
access only the compositional information yet make implicit assumptions
about the scale.

Recently, Nixon et al. (8) showed that the challenge of non-biological
variation in sequencing depth be viewed as a problem of
partially-identified models. They showed that \emph{all} normalizations
make some assumption about scale. These implicit assumptions are often
difficult to interpret, and different normalizations provide different
outputs when applied to the same dataset (3, 21--24). Intuitively,
normalizations in widespread use assume that either all samples have the
same scale, e.g.~proportions, rarefaction (25), RPKM (26, 27), etc; or
that a subset of features in one sample can be chosen as a reference to
which the others are scaled e.g.~the TMM (28), or LVHA (7) or the
additive log-ratio (29); or that different sub-parts of each sample
maintain a constant scale across samples e.g.~the RLE (30); or that the
geometric mean of the parts is appropriate e.g the CLR and its
derivatives.

The original ALDEx2 (31) model unwittingly made a strict assumption
about scale through the CLR normalization (8). While this assumption
could be useful in many cases it could never be exactly true, and others
have shown that it is not always the best option (32). Nixon et al. (8)
showed that better scale assumptions resulted in more reproducible data
analysis including better control of both false positive and false
negative results. In essence ALDEx2 has been modified to explicitly
model the scale over a range of reasonable normalization parameters.
Here, we briefly introduce these modifications and then show how scale
uncertainty can greatly improve modeling in transcriptome and
meta-transcriptome datasets to provide more robust and reproducible
results.

\section{Implementation}\label{implementation}

To be concrete, we let \(\mathbf{Y}\) denote the \emph{measured}
\(D \times N\) matrix of sequence counts with elements
\(\mathbf{Y}_{dn}\) indicating the number of measured DNA molecules
mapping to feature \(d\) (e.g., a taxon or gene) in sample \(n\).
Likewise, we can denote \(\mathbf{W}\) as the \emph{true} amount of
class \(d\) in the biological system from which sample \(n\) was
obtained. We can think of \(\mathbf{W}\) as consisting of two parts, the
scale \(\mathbf{W}^{\perp}\) (e.g., totals) and the composition
\(\mathbf{W}^{\parallel}\) (i.e., proportions). That is,
\(\mathbf{W}^{\perp}\) is a \(N\)-vector with elements
\(\mathbf{W}^{\perp}_{n}=\sum_{d}\mathbf{W}_{dn}\) while
\(\mathbf{W}^{\parallel}\) is a \(D \times N\) matrix with elements
\(\mathbf{W}^{\parallel}_{dn}=\mathbf{W}_{dn}/\mathbf{W}^{\perp}_{n}\).
Note that with these definitions \(\mathbf{W}\) can be written as the
element-wise combination of scale and composition:
\(\mathbf{W}_{dn}=\mathbf{W}^{\parallel}_{dn}\mathbf{W}^{\perp}_{n}\).

All the normalizations in current use are data transformations that can
be stated as ratios of the form
\(\hat{{\mathbf{W}}}_{dn}=\mathbf{Y}_{dn}/f(\mathbf{Y})\), where the
denominator is determined by some function of the observation. We use
the \(\hat{{\mathbf{W}}}\) notation to indicate that the normalization
is attempting to provide an estimate of the true value. The technical
variation in sequencing depth
(\(\mathbf{Y}^{\perp}_{n}=\sum_{d}\mathbf{Y}_{dn}\)) implies that
observed data \(\mathbf{Y}\) provides us with information about the
system composition \(\mathbf{W}^{\parallel}\) but little to no
information in the system scale \(\mathbf{W}^{\perp}\) (Lovell et
al.~2011).

\subsection{Adding Scale Uncertainty in
ALDEx2}\label{adding-scale-uncertainty-in-aldex2}

The ALDEx2 R package (31, 33) is a general purpose toolbox for Bayesian
modeling of HTS data. For brevity, we discuss ALDEx2 in its simplest
form as a tool for estimating the magnitude and statistical significance
of Log-Fold-Changes (LFC) (e.g., differential abundance or differential
expression analysis), but note that it can be used to fit more complex
linear models. At a high-level, ALDEx2 involves three steps: 1) a
Bayesian model is used to estimate \(\mathbf{\hat{W}}^{\parallel}\)
given observations \(\mathbf{Y}\); 2) a normalization is used to
estimate \(\mathbf{\hat{W}}\) given the estimate
\(\mathbf{\hat{W}}^{\parallel}\); 3) the \(\mathbf{\hat{W}}\) estimates
are used to estimate LFC as part of differential abundance (or
expression) analyses. For more details on ALDEx2 see (8, 31).

By default, ALDEx2 uses the CLR normalization which can be written as
\(\log \mathbf{\hat{W}}_{dn}=\log (\mathbf{\hat{W}}^{\parallel}_{dn}/G_{n})\)
where \(G_{n}\) denotes the geometric mean of the vector
\((\hat{W}^{\parallel}_{1n}, \dots, \hat{W}^{\parallel}_{Dn})\). The CLR
normalization equates to an implicit assumption that
\(\mathbf{\hat{W}}^{\perp}_{n}=1/G_{n}\)(8) and even slight errors in
this assumption introduces bias into the LFC estimate that is not
accounted for when estimating uncertainty (e.g., confidence intervals or
credible sets). In fact, Nixon et al. (8) showed that the only way in
which the ALDEx2 model, or any normalization-based model, could ever be
calibrated (e.g., control Type-I Error rates) was if this assumption was
exactly true. In support of this assertion it is known that false
discovery rates vary widely by analysis, dataset, and normalization
method (34, 35). To fix this, Nixon et al. (8) showed that models should
incorporate potential error in the assumptions implied by
normalizations.

Nixon et al.~(2023) generalized the concept of normalizations by
introducing the concept of a \textit{scale model} to account for
potential error. Scale models can be incorporated into ALDEx2, turning
the ALDEx2 model into a specialized type of statistical model which they
called a \textit{Scale Simulation Random   Variable} (SSRV). They did
this by including a model for \(\mathbf{\hat{W}}^{\perp}_{n}\). Since
the CLR normalization makes the assumption
\(\mathbf{\hat{W}}^{\perp}_{n}=1/G_{n}\), the CLR normalization can be
generalized by considering probability models for the scale
\(\mathbf{\hat{W}}^{\perp}_{n}\) that have mean \(1/G_{n}\). For
example, the following scale model generalizes the CLR:

\[\log \mathbf{\hat{W}}^{\perp}_{n} = -\log G_{n} + \Lambda x_{n} \qquad \Lambda \sim N(0, \gamma^{2})\]

where \(\gamma\) is a tunable parameter drawn from a Gaussian
distribution that controls the degree of uncertainty in the CLR
assumption and \(x_{n}\) denotes a binary condition indicator (e.g.,
\(x_{n}=1\) denotes case and \(x_{n}=0\) denotes control). We have made
those modifications a permanent fixture of ALDEx2 which now represents
the first software package designed for SSRV-based inference.

\section{Results}\label{results}

\subsection{Adding scale uncertainty replaces the need for dual
significance
cutoffs.}\label{adding-scale-uncertainty-replaces-the-need-for-dual-significance-cutoffs.}

It is standard practice in many fields of HTS, but especially
transcriptomics, to use the dual cutoff approach graphically exemplified
by volcano plots (36, 37) because not all statistically significant
differences are biologically relevant. Paraphrasing this we can say that
in in some types of datasets more samples leads to a majority of
features being statistically significant. A detailed reasoning for this
using modelled data is explained in two reports by Nixon et al. (8, 9),
which shows that scaled models completely address this analytic issue.

We use the data of Gierliński et al. (38) who conducted a highly
replicated yeast transcriptome experiment comparing a wild-type strain
with a snf2 gene knockout, \(\Delta\)snf2. This dataset has been used to
argue that a dual cutoff approach is appropriate to limit the number of
significant parts and for purposes of reproducibility (37). However, in
this benchmarking study, the number of significantly different
transcripts varied between 65\% to \textgreater80\% of all transcripts
depending on the tool; in essence the desired behavior was that almost
all transcripts were significantly different. The guidance on
reproducibility runs counter to standard statistical practice where
power is intrinsically linked to sample size and very large sample sizes
are indeed desired (39). Furthermore, the dual-cutoff approach is known
not to provide appropriate FDR control (5, 6). Schurch et al. (37)
further suggested that different tools might be better in some
conditions or datasets than others because each tool has different
intrinsic statistical power and Type 1 and Type 2 errors. Through the
lens of scale uncertainty, the behaviour of the tools in this study show
unacknowledged bias; false confidence in the precision of the estimate
as sample size increases because o mis-specification in the scale of the
data. This certainty is driven by the assumptions of the tool not the
actual experiment being investigated (9).

Using either DESeq2 or ALDEx2, a majority of transcripts are
statistically significantly different between groups with a
Benjamini-Hochberg (40) false discovery rate (FDR) of 0.05; i.e.~4636
(79\%, DESeq2) or 4172 (71\%, ALDEx2) of the 5891 transcripts. Such
large numbers of significant transcripts seems biologically unrealistic
and furthermore breaks the necessary assumption made by the
normalization methods that the majority of the features must be
invariant. That 118 transcripts are identified by ALDEx2 and not DESeq2,
while DESeq2 identifies 582 transcripts that ALDEx2 does not, suggests
that the choice of normalization plays a role in which results are
returned as significant and that some if not the majority are driven by
technical differences in the analysis (22, 23, 37, 41, 42).

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{go3_files/figure-latex/plot1-1.pdf}}
\caption{Effect and volcano plots for unscaled and scaled transcriptome
analysis. ALDEx2 was used to conduct a differential abundance (DA)
analysis on the yeast transcriptome dataset. The results were plotted to
show the relationship between difference and dispersion using effect
plots ) or difference and the Benjamini-Hochberg corrected p-values
(volcano plot). Panels A,C are for the unscaled analysis, and Panels B,D
are for the scaled analysis. Each point represents the values for one
transcript, with the color indicating if that transcript was significant
in the scaled analysis and unscaled analysis (red) or in the unscaled
analysis only (orange). Points in grey are not statistically
signficantly different with any analysis. The horizontal dashed lines
represent a log2(difference) of \(\pm 1.4\).}
\end{figure}

The effect plots (43) in Figure 1A (ALDEx2) and Supplementary Figure 1
(DESeq2) shows that the majority of significant transcripts (red,
orange) have negligible differences between groups and very low
dispersion and we suggest that this is driven by the experimental design
(37). Scale uncertainty can be incorporated using the \texttt{gamma}
parameter that controls the amount uncertainty added to the CLR mean
assumption when we call either \texttt{aldex()}, or
\texttt{aldex.clr()}. Figure 1B shows that setting \(\gamma=0.5\)
results in far fewer transcripts being significant (205) and we observe
that the minimum dispersion increases from 0.12 (unscaled) to 0.67
(scaled). The Volcano plots in Figure 1 C and D show a similar story.
Here we can see that adding scale increases the minimum FDR value and
increases the concordance between the FDR value and the difference
between groups (compare panels C and D).

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{go3_files/figure-latex/disp-1.pdf}}
\caption{Adding scale uncertainty changes the dispersion distribution.
Panel A shows a plot of the expected value for relative abundance vs the
expected value for the pooled dispersion as output by
\texttt{aldex.effect}. The dashed horizontal lines show the median value
for the features with a rAbundance between -0.5 and 0.5, and the light
colored lines are lowess lines of fit through the center of mass of the
data. Panel B plots the dispersion difference between \(\gamma = 1\) and
\(\gamma = 0\); note the non-linear relationship that highlights the
rotation that is evident in Panel A. The colored lines indicate the
lowess line of fit through the centre of mass of the plot for the
various populations of points. The grey line is the total population and
shows the difference \(\Delta\), the red line is the population of
significant transcripts (*) with no scale, the orange line is the
population of significant transcripts with a difference threshold (T) of
about \(\pm 2^{1.4}\)-fold change, the blue line is the population of
significant transcripts with \(\gamma = 0.5\), and the cyan line is the
significant population with \(\gamma = 1\). \(\Delta\): Difference, *:
significant, T: thresholded.}
\end{figure}

As shown by the effect plot in Figure 1 the root cause of the many
statistically significant positive transcripts is the very large number
of transcripts with negligible variance. This phenomenon is not unique
to ALDEx2 as Supplementary Figure 1 shows that the same phenomenon
occurs in DESeq2 (and presumably other methods although the relevant
parameters are not exposed). This issue is not unique to this dataset
and it is common practice to use a dual-cutoff by choosing transcripts
based on a thresholds for both corrected p-values and fold-changes (37)
(here set at \(\pm 2^{1.4}\) for the latter), although considerable
variation in cutoff values is observed. These limits are shown by the
dashed grey lines. Here, applying a dual-cutoff using a heuristic of at
least a \(2^{1.4}\) fold change reduces the number of significant
outputs to 193 for DESeq2 and to 186 for ALDEx2, and is in-line with
that observed with that found by ALDEx2 with \(\gamma = 0.5\) which
identifies 205. Indeed, Supplementary Figure 2, shows that even adding a
very small amount of scale \(\gamma = 0.1\) reduces the number of
significant transcripts by more than half and the
\texttt{aldex.scaleSim()} function can be used to identify those
transcripts that are significant only because of an absence of scale.
These results begs the question: why bother with significance tests at
all if all transcripts with \(> 1.4\)-fold expression change are
statistically significant?

The effect on dispersion with increasing amounts of scale are shown in
Figure 2A, where we can see that the dispersion increases as scale is
added. Note that the dispersion in the unscaled data in Figure 2A
reaches a minimum near the mid-point of the distribution, and also does
so when the analysis is conducted with DESeq2 (Supplementary Figure 3).
This shows more clearly that dispersion of many transcripts is almost
negligible in the absence of scale and makes the counter-intuitive
suggestion that the variance in expression of the majority of genes with
moderate expression is more predictable than highly-expressed genes or
of housekeeping genes (44). This is at odds with the known biology of
cells where single cell counting of highly-expressed transcripts shows
that they have little intrinsic variation (45).
